Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2021-08-27 fstats_pca_lewontin.tex
revision part 1

In most applications, we do not use all PCs, but instead truncate to the first $K$ PCs, which explain most of the between-population genetic variation.
Thus, 
\begin{align}
F_2(P_i, P_j) &=& \sum_{k=1}^K(p_{ik} - p_{jk})^2 &+& \sum_{k=K+1}^n(p_{ik} - p_{jk})^2&&&&\nonumber\\
&=& {\hat{F_2}^{(K)}(P_i, P_j)} &+& {\epsilon^{(K)}(P_i, P_j)}&&&& \text{.}
\end{align}
In this notation, $\hat{F_2}^{(K)}$ is the approximation of $F_2$ with only the first $K$ PCs considered, and $\epsilon^{(K)}$ is the corresponding approximation error.
If we sum up the squared approximation errors over all pairs of populations, we obtain 
\begin{equation}
\sum_{i,j} \epsilon^{(K)}(P_i, P_j)^2 = \sum_{i,j} \left(\hat{F_2}^{(K)}(P_i, P_j) - F_2^{(K)}(P_i, P_j)\right)^2 = \normsq{\MF - \hat{\MF}}_F \text{,}
\end{equation}
where the Frobenius-norm $\normsq{\cdot}_F$ of a matrix is defined as the square root of the sum-of-squares of all its elements. This is precisely the function that is minimized in MDS \citep{jolliffe2013}. In that sense, $\hat\MF^{(K)}$ is the optimal low-rank approximation of $\MF$ for any $K$ in that it minimizes the approximation error of all $F_2$-statistics.

@@

\subsection{$F_2$ in PC-space}
The $F_2$-statistic is an estimate of the squared Euclidean distance between two
populations. It thus corresponds to the squared distance between populations in PCA-space, and
reflects the intuition that closely related populations  will be close to each other on a
PCA-plot, and have low pairwise $F_2$-statistics. In converse, if two
populations have high $F_2$ but appear on the same point on an PCA-plot, this suggests
that substantial variation is hidden on higher PCs.

>>>>

In most applications, we do not use all PCs, but instead truncate to the first $K$ PCs, which explain most of the between-population genetic variation.
Thus, 
\begin{align}
F_2(P_i, P_j) &=& \sum_{k=1}^K(p_{ik} - p_{jk})^2 &+& \sum_{k=K+1}^n(p_{ik} - p_{jk})^2&&&&\nonumber\\
&=& {\hat{F_2}^{(K)}(P_i, P_j)} &+& {\epsilon^{(K)}(P_i, P_j)}&&&& \text{.}
\end{align}
In this notation, $\hat{F_2}^{(K)}$ is the approximation of $F_2$ with only the first $K$ PCs considered, and $\epsilon^{(K)}$ is the corresponding approximation error. I will omit the superscript when the exact number of PCs is not relevant.
If we sum up the squared approximation errors over all pairs of populations, we obtain 
\begin{equation}
\sum_{i,j} \epsilon^{(K)}(P_i, P_j)^2 = \sum_{i,j} \left(\hat{F_2}^{(K)}(P_i, P_j) - F_2^{(K)}(P_i, P_j)\right)^2 = \normsq{\MF - \hat{\MF}}_F \text{,}
\end{equation}
where the Frobenius-norm $\normsq{\cdot}_F$ of a matrix is defined as the square root of the sum-of-squares of all its elements. This is precisely the function that is minimized in MDS \citep{jolliffe2013}. In that sense, $\hat\MF^{(K)}$ is the optimal low-rank approximation of $\MF$ for any $K$ in that it minimizes the approximation error of all $F_2$-statistics.

@@

\subsection{$F_2$ in PC-space}
The $F_2$-statistic is an estimate of the squared Euclidean distance between two
populations. It thus corresponds to the squared distance between populations in PCA-space, and
reflects the intuition that closely related populations  will be close to each other on a
PCA-plot, and have low pairwise $F_2$-statistics. In converse, if two
populations have high $F_2$ but appear on the same point on an PCA-plot, this suggests
that substantial variation is hidden on higher PCs, and we may want to investigate the PCs which contribute a large distance to this particular $F_2$-statistic in order to better understand and visualize their relationship.
